Indecision under time pressure arises from suboptimal switching behaviour

### 2024-10-03 paper/Cover_letter_Indecisions.tex
submitted to jep general

\begin{document}
\fontsize{12pt}{12pt}\selectfont
%\pagenumbering{gobble}
\pagestyle{fancy}
\cfoot{}
\lfoot{}
\rfoot{\textcolor{myudblue}{1}}
\raggedright
\begin{minipage}[ht]{0.6\textwidth}
\hspace{-10mm}\includegraphics[height=0.9in]{udcoe_logo}
\end{minipage}\hfill
\begin{minipage}[ht]{0.275\textwidth}
\begin{tabular}{lr}
\raggedleft
&\fontsize{10}{10}\selectfont \hspace{+20mm}\textcolor{myudblue}{Joshua Cashaback}\\
&\fontsize{10}{10}\selectfont \textcolor{myudblue}{Assistant Professor}\\
 &{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\faPhone \hspace{1mm} +1.302.784.5986}}\\
 &{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\href{mailto:joshcash@udel.edu}{\faEnvelope \hspace{1mm} joshcash@udel.edu}}}\\
%&{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\faHome joshcashaback.weebly.com}}
%&{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\href{mailto:cashabackjga@gmail.com}{cashabackjga@gmail.com}}}\\
\end{tabular}
\end{minipage}
\\
\vspace{6mm}
\today
\\
\vspace{6mm}
Dear Editor,
\justify
\vspace{1.0mm}
We are writing to request that you consider our paper, \textcolor{myudblue}{\textbf{Indecision under time pressure arises from suboptimal switching behaviour}}, for publication in \emph{PLOS Computational Biology}.
\vspace{3.0mm}
\\
Indecisions can be catastrophic, preventing a life-saving response or leading to a loss of large sums of money.  Despite their importance and high prevalence, indecisions have been surprisingly ignored in the literature. The main culprit is that the field of decision-making leans heavily on the classic two-alternative forced choice task---where a human or animal \emph{must} select one of two potential options---which simply does not permit a non-response past some deadline (i.e., indecision). Of the few papers that have considered a deadline, indecisions are usually only mentioned in passing and, more often, simply removed from analyses and discarded from model fitting. \textbf{The field has consequently overlooked a common and hallmark feature of human and animal behaviour---indecisions.}
\vspace{3mm}
\\
Here we designed a novel paradigm to study high pressure scenarios that lead to indecisions. We found a median indecision rate of approximately 20\%, \textbf{with some participants even displaying over 50\% indecisions despite having plenty of time to make a guess.} Our computational modelling suggested that these indecisions arose by not accounting for the time delay and uncertainty associated when switching from reacting to guessing. In a second experiment, we show for the first time there is an inherent and functionally relevant time delay when switching from reacting to guessing, a phenomenon previously unreported in the literature. \textbf{Our paper novelly provides a mechanism to explain how indecisions arise.}
\vspace{3mm}
\\
We fully expect that our work will broadly draw interest from fields as diverse as Decision-Making, Economics, and Sports Science, which are all interested in the role of decision-making on behaviour.
Thank you for your time and consideration.
\vspace{3mm}
\\
Sincerely, 
\vspace{3mm}
\\
\includegraphics[height=0.75in]{signature}
\\
\vspace{-3.5mm}
\\
Joshua G. A. Cashaback, Ph.D.
\thispagestyle{fancy}
\rfoot{\textcolor{myudblue}{\thepage}}
\lfoot{\textcolor{myudblue}{\textbf{\fontsize{10}{10}\selectfont Biomedical Engineering | College of Engineering | University of Delaware} }
\\
\fontsize{10}{10}\selectfont \textcolor{myudblue}{STAR Health Sciences Complex, Rm 201J | 540 South College Avenue | Newark, DE., 19713}}
\newpage
\noindent \textcolor{myudblue}{REVIEWER SUGGESTIONS}
\\
\\
Michael S Landy, New York University, landy@nyu.edu
\\
\\
Han L.J. van der Maas, University of Amsterdam, H.L.J.vanderMaas@uva.nl 
\\
\\
Tony Carlsen, University of Ottawa, acarlsen@uottawa.ca
\\
\\
Martin Wiener, George Mason University, mwiener@gmu.edu
%\\
%\\
%Laurence T. Maloney, New York University, laurence.maloney@nyu.edu
%\\
%\\
%Jan Drugowitsch, Harvard Medical School, jan_drugowitsch@hms.harvard.edu
%\\
%\\
%Taraz G. Lee, University of Michigan, tarazlee@umich.edu
%\\
%\\
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%% End CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\begin{document}
\fontsize{12pt}{12pt}\selectfont
%\pagenumbering{gobble}
\pagestyle{fancy}
\cfoot{}
\lfoot{}
\rfoot{\textcolor{myudblue}{1}}
\raggedright
\begin{minipage}[ht]{0.6\textwidth}
\hspace{-10mm}\includegraphics[height=0.9in]{udcoe_logo}
\end{minipage}\hfill
\begin{minipage}[ht]{0.275\textwidth}
\begin{tabular}{lr}
\raggedleft
&\fontsize{10}{10}\selectfont \hspace{+20mm}\textcolor{myudblue}{Joshua Cashaback}\\
&\fontsize{10}{10}\selectfont \textcolor{myudblue}{Assistant Professor}\\
 &{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\faPhone \hspace{1mm} +1.302.784.5986}}\\
 &{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\href{mailto:joshcash@udel.edu}{\faEnvelope \hspace{1mm} joshcash@udel.edu}}}\\
%&{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\faHome joshcashaback.weebly.com}}
%&{\fontsize{10}{10}\selectfont \textcolor{myudblue}{\href{mailto:cashabackjga@gmail.com}{cashabackjga@gmail.com}}}\\
\end{tabular}
\end{minipage}
\\
\vspace{6mm}
\today
\\
\vspace{6mm}
Dear Editor,
\justify
\vspace{1.0mm}
We are writing to request that you consider our paper, \textcolor{myudblue}{\textbf{Indecision under time pressure arises from suboptimal switching behaviour}}, for publication in \emph{Journal of Experimental Psychology: General}.
\vspace{3.0mm}
\\
Indecisions can be catastrophic, preventing a life-saving response or leading to a loss of large sums of money.  Despite their importance and high prevalence, indecisions have been surprisingly ignored in the literature. The main culprit is that the field of decision-making leans heavily on the classic two-alternative forced choice task---where a human or animal \emph{must} select one of two potential options---which simply does not permit a non-response past some deadline (i.e., indecision). Of the few papers that have considered a deadline, indecisions are usually only mentioned in passing and, more often, simply removed from analyses and discarded from model fitting. \textbf{The field has consequently overlooked a common and hallmark feature of human and animal behaviour---indecisions.}
\vspace{3mm}
\\
Here we designed a novel paradigm to study high pressure scenarios that lead to indecisions. We found a median indecision rate of approximately 20\%, \textbf{with some participants even displaying over 50\% indecisions despite having plenty of time to make a guess.} Our computational modelling suggested that these indecisions arose by not accounting for the time delay and uncertainty associated when switching from reacting to guessing. In a second experiment, we show for the first time there is an inherent and functionally relevant time delay when switching from reacting to guessing, a phenomenon previously unreported in the literature. \textbf{Our paper novelly provides a mechanism to explain how indecisions arise.}
\vspace{3mm}
\\
We fully expect that our work will broadly draw interest from fields as diverse as Decision-Making, Economics, and Sports Science, which are all interested in the role of decision-making on behaviour.
Thank you for your time and consideration.
\vspace{3mm}
\\
Sincerely, 
\vspace{3mm}
\\
\includegraphics[height=0.75in]{signature}
\\
\vspace{-3.5mm}
\\
Joshua G. A. Cashaback, Ph.D.
\thispagestyle{fancy}
\rfoot{\textcolor{myudblue}{\thepage}}
\lfoot{\textcolor{myudblue}{\textbf{\fontsize{10}{10}\selectfont Biomedical Engineering | College of Engineering | University of Delaware} }
\\
\fontsize{10}{10}\selectfont \textcolor{myudblue}{STAR Health Sciences Complex, Rm 201J | 540 South College Avenue | Newark, DE., 19713}}
\newpage
\noindent \textcolor{myudblue}{REVIEWER SUGGESTIONS}
\\
\\
Laurence Maloney, New York University, laurence.maloney@nyu.edu

Expert in statistical decision theory and computational modeling of decisions
\\
\\
Michael S Landy, New York University, landy@nyu.edu

Expert in statistical decision theory and computational modeling of decisions
\\
\\
Heather Neyedli, Dalhouse University, hneyedli@dal.ca

Expert in decision-making under risk and optimal decision-making
\\
\\
Todd Hudson, New York University, Todd.Hudson@nyulangone.org

Expert in optimal timing and decision-making
%\\
%\\
%Laurence T. Maloney, New York University, laurence.maloney@nyu.edu
%\\
%\\
%Jan Drugowitsch, Harvard Medical School, jan_drugowitsch@hms.harvard.edu
%\\
%\\
%Taraz G. Lee, University of Michigan, tarazlee@umich.edu
%\\
%\\
\\
\\
\noindent \textcolor{myudblue}{EDITOR SUGGESTIONS}
\\
\\
Joo-Hyun Song
\\
\\
Jared Medina  
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%% End CV Document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-10-03 paper_apa/main.tex
submitted to jep general

\newpage
% \pagenumbering{arabic} % page numbering starts here
\vspace{2mm}
\section{Introduction}
\vspace{2mm}
\noindent Indecisions arise from failing to decide and act upon sensory information in time, such as a driver failing to brake or hit the gas pedal when a traffic light turns yellow. When acting under high pressure time constraints, the ability to accurately time a decision is critical to success. The vast majority of decision-making research either does not consider responses made after some time constraint or simply does not permit a non-response, such as in the classic two-alternative forced choice paradigm \autocite{zacksenhouseRobustOptimalStrategies2010, bogaczPhysicsOptimalDecision2006, choMechanismsUnderlyingDependencies2002, joganNewTwoalternativeForced2014, ratcliffDiffusionDecisionModel2008,ulrichThresholdEstimationTwoalternative2004,tylerSignalDetectionTheory2000}. Thus, despite its real-world ubiquity and importance, we have very little understanding of how indecisions arise.

@@

The incorrect percentage across all conditions was 26.3\% [range: 0.0\%, 57.5\%; \boldblue{Fig. 2E}]. We found significant interactions between stimulus movement onset mean and variance for incorrect decisions (F[1.66,31.51] = 3.72, p = 0.033). Participants displayed a greater percentage of incorrect decisions in the \emph{late mean} condition than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 92.5\%). 
For indecisions, wins, and incorrects we also found the same significant differences between conditions when separately analyzing the first half and second half of trials (\boldblue{Supplementary C}). 
That is, the same trends for the first and second half of trials shows that participants determined their decision-timing strategy early on in each condition and there was a negligible influence of learning.

@@

\vspace{5mm}
\subsection{Experiment 1 Design}

@@

\vspace{2mm}
\subsection{Experiment 2 Design}

@@

\vspace{2mm}
\subsection{Data Analysis}

@@

% \newpage
% \input{supplementary}

>>>>

\newpage
% \pagenumbering{arabic} % page numbering starts here
Word count: 7114
\section{Introduction}
\vspace{2mm}
\noindent Indecisions arise from failing to decide and act upon sensory information in time, such as a driver failing to brake or hit the gas pedal when a traffic light turns yellow. When acting under high pressure time constraints, the ability to accurately time a decision is critical to success. The vast majority of decision-making research either does not consider responses made after some time constraint or simply does not permit a non-response, such as in the classic two-alternative forced choice paradigm \autocite{zacksenhouseRobustOptimalStrategies2010, bogaczPhysicsOptimalDecision2006, choMechanismsUnderlyingDependencies2002, joganNewTwoalternativeForced2014, ratcliffDiffusionDecisionModel2008,ulrichThresholdEstimationTwoalternative2004,tylerSignalDetectionTheory2000}. Thus, despite its real-world ubiquity and importance, we have very little understanding of how indecisions arise.

@@

The incorrect percentage across all conditions was 26.3\% [range: 0.0\%, 57.5\%; \boldblue{Fig. 2E}]. We found significant interactions between stimulus movement onset mean and variance for incorrect decisions (F[1.66,31.51] = 3.72, p = 0.033). Participants displayed a greater percentage of incorrect decisions in the \emph{late mean} condition than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 92.5\%). 
For indecisions, wins, and incorrects we also found the same significant differences between conditions when separately analyzing the first half and second half of trials (\boldblue{Supplementary B}). 
That is, the same trends for the first and second half of trials shows that participants determined their decision-timing strategy early on in each condition and there was a negligible influence of learning.

@@

\subsection{Experiment 1 Design}

@@

\subsection{Experiment 2 Design}

@@

\subsection{Data Analysis}

@@

\newpage
\input{supplementary}

### 2024-10-03 paper_apa/supplementary.tex
apa formatted
submitted to jep general

% \definecolor{myudblue}{RGB}{0, 83, 159}
\begin{document}
\SectionHeader{Supplementary}

\noindent\boldblue{\large{Supplementary A: Low and High Variance Results}}
\begin{figure}[H]
    \centering
    \includegraphics[scale =1]{figures/exp1_movement_onset_panel_all_conditions.png}
    \caption*{\boldblue{Supplementary Figure 1: Timing Behaviour.} \boldblue{A)} Average participant movement onset (y-axis) for each experimental condition (x-axis). We found a significant main effect of stimulus movement onset mean (F[1.55,29.48] = 4.36, p = 0.030) and variance (F[1.00,19.00], p = 0.017)) on participant movement onset. There was no significant interaction (F[1.66,31.47], p = 0.565) \boldblue{B)} Corresponding participant movement onset (y-axis) for the three different mean stimulus onsets (x-axis) when collapsed across variance. During the early mean conditions, participants had a high probability of reacting to the stimulus and reaching a target within the time constraint. For the middle mean conditions, participants often waited longer to react and had a later guess than the early mean conditions. Waiting longer to react and/or late guessing led to later movement onset. In the late mean condition, participants waited to react to the stimulus, but ended up guessing a majority of the time (see \boldblue{Supplementary Figure 6}). The combination of waiting for the stimulus and eventually guessing, led to an earlier movement onset compared to the middle mean condition. For both the middle mean and late mean conditions, late guessing often led to indecisions (see \boldblue{Fig. 2}). \boldblue{C)} Participant movement onset standard deviation (y-axis) for each experimental condition (x-axis). We found a significant main effect of mean (F[1.38, 26.28], p = 0.018) and variance (F[1.00,19.00], p < 0.001) on participant movement onset standard deviation There was no significant interaction between stimulus movement onset mean and variance (F[1.98, 37.58], p = 0.097) \boldblue{D)} Corresponding participant movement onset standard deviation (y-axis) for the three different mean onsets (x-axis), when collapsed across variance. In the late mean conditions, waiting to react and then guessing led to a larger movement onset standard deviation compared to the early mean and middle mean conditions. Box and whisker plots display the 25th, 50th, and 75th percentiles. Open circles and connecting lines represent individual data.}
\end{figure}

@@

\newpage
\noindent\boldblue{\large{Supplementary B: First Half and Second Half Data Trends}}
\vspace{-1mm}

@@

\begin{figure}[H]
\begin{minipage}[c]{0.6\textwidth}
    \includegraphics[scale = 1]{figures/exp1_data_panel_second_half.png}
\end{minipage}\hfill
\begin{minipage}[c]{0.3\textwidth}
    \caption*{
        \boldblue{Supplementary Figure 4: Outcome Metrics for Second Half of Trials} \boldblue{A)} Indecisions (\%) \boldblue{B)} Wins (\%), and \boldblue{C)} Incorrects (\%) for the second half of trials. 
        Here we show that the significant differences between conditions are the same for the first half (\boldblue{Supplementary Figure 3}) and second half of trials.         
        The same trends for the first half and second half of trials indicates that participants determined their timing strategy early on in each condition and there was a negligible effect of learning. 
    } 
\end{minipage}
\end{figure}

\newpage
\noindent\boldblue{\large{Supplementary C: Optimal Models}}
\vspace{-1mm}

@@

\vspace{2mm}
\noindent \boldblue{General Formulation for the Models}

@@

\vspace{2mm}
\noindent\boldblue{Model Parameter Estimation and Fitting Procedure}

@@

\vspace{2mm}
\noindent \boldblue{Model Parameter Values}

@@

\newpage
\noindent\boldblue{\large\textcolor{mydarkblue}{Supplementary D: Model Results}}

@@

\newpage
\noindent\boldblue{\large\textcolor{mydarkblue}{Supplementary E: Response Time Task}}

@@

    \caption*{\boldblue{Supplementary Figure 8: Response Time Experiment}. One of two potential targets (dashed white rings; not visible to participants) would appear on the screen. Once a target appeared (solid white ring), participants were instructed to rapidly move their cursor to hit the displayed target. We used the mean and standard deviation of response times as model parameters in the optimal transition models.}
\end{figure}

\newpage
\noindent\boldblue{\large\textcolor{mydarkblue}{Supplementary F: Timing Uncertainty Task}}

@@

    \caption*{\boldblue{Supplementary Figure 9: Timing Uncertainty Experiment}. One of two potential targets (dashed white rings; not visible to participants) would appear on the screen. Once a target appeared (solid white ring), participants were instructed to rapidly move their cursor to hit the displayed target. We used the mean and standard deviation of response times as model parameters in the optimal transition models.}
\end{figure}

\newpage
\noindent\boldblue{\large\textcolor{mydarkblue}{Supplementary G: Guess Decisions}}

@@

    \caption*{\boldblue{Supplementary Figure 10: Guess Decision Percentages for Experiment 1}. 
    Estimated percentage of guess decision percentages (y-axis) for each experimental condition (x-axis). During \boldblue{Experiment 1} participants were rewarded for successful trials and experienced a time constraint. Thus, we would expect participants to have faster response times in \boldblue{Experiment 1} than in the Response Time task \autocite*{milsteinInfluenceExpectedValue2007,shadmehrMovementVigorReflection2019a}. To account for this, we adjusted participant response times by subtracting 25ms from their median response time in the Response Time task. A trial was a guess decision if the difference between the stimulus movement onset and participant movement onset was less than that participant’s adjusted response time. We found that participants guessed significantly more often as the stimulus’s mean movement onset approached the time constraint.}
\end{figure}

>>>>

% \definecolor{myudblue}{RGB}{0, 83, 159}
\begin{document}
\section{Supplementary}

\section{Supplementary A: Low and High Variance Results}
\begin{figure}[H]
    \centering
    \includegraphics[scale =1]{figures/exp1_movement_onset_panel_all_conditions.png}
    \caption*{\boldblue{Supplementary Figure 1: Timing Behaviour.} \boldblue{A)} Average participant movement onset (y-axis) for each experimental condition (x-axis). We found a significant main effect of stimulus movement onset mean (F[1.55,29.48] = 4.36, p = 0.030) and variance (F[1.00,19.00], p = 0.017)) on participant movement onset. There was no significant interaction (F[1.66,31.47], p = 0.565) \boldblue{B)} Corresponding participant movement onset (y-axis) for the three different mean stimulus onsets (x-axis) when collapsed across variance. During the early mean conditions, participants had a high probability of reacting to the stimulus and reaching a target within the time constraint. For the middle mean conditions, participants often waited longer to react and had a later guess than the early mean conditions. Waiting longer to react and/or late guessing led to later movement onset. In the late mean condition, participants waited to react to the stimulus, but ended up guessing a majority of the time (see \boldblue{Supplementary Figure 6}). The combination of waiting for the stimulus and eventually guessing, led to an earlier movement onset compared to the middle mean condition. For both the middle mean and late mean conditions, late guessing often led to indecisions (see \boldblue{Fig. 2}). \boldblue{C)} Participant movement onset standard deviation (y-axis) for each experimental condition (x-axis). We found a significant main effect of mean (F[1.38, 26.28], p = 0.018) and variance (F[1.00,19.00], p < 0.001) on participant movement onset standard deviation There was no significant interaction between stimulus movement onset mean and variance (F[1.98, 37.58], p = 0.097) \boldblue{D)} Corresponding participant movement onset standard deviation (y-axis) for the three different mean onsets (x-axis), when collapsed across variance. In the late mean conditions, waiting to react and then guessing led to a larger movement onset standard deviation compared to the early mean and middle mean conditions. Box and whisker plots display the 25th, 50th, and 75th percentiles. Open circles and connecting lines represent individual data.}
\end{figure}

@@

\newpage
\section{Supplementary B: First Half and Second Half Data Trends}
\vspace{-1mm}

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/exp1_data_panel_second_half.png}

    \caption*{
        \boldblue{Supplementary Figure 4: Outcome Metrics for Second Half of Trials} \boldblue{A)} Indecisions (\%) \boldblue{B)} Wins (\%), and \boldblue{C)} Incorrects (\%) for the second half of trials. 
        Here we show that the significant differences between conditions are the same for the first half (\boldblue{Supplementary Figure 3}) and second half of trials.         
        The same trends for the first half and second half of trials indicates that participants determined their timing strategy early on in each condition and there was a negligible effect of learning. 
    } 
\end{figure}

\newpage
\section{Supplementary C: Optimal Models}
\vspace{-1mm}

@@

\vspace{2mm}
\subsection{General Formulation for the Models}

@@

\vspace{2mm}
\subsection{Model Parameter Estimation and Fitting Procedure}

@@

\vspace{2mm}
\subsection{Model Parameter Values}

@@

\newpage
\section{Supplementary D: Model Results}

@@

\newpage
\section{Supplementary E: Response Time Task}

@@

    \caption*{\boldblue{Supplementary Figure 7: Response Time Experiment}. One of two potential targets (dashed white rings; not visible to participants) would appear on the screen. Once a target appeared (solid white ring), participants were instructed to rapidly move their cursor to hit the displayed target. We used the mean and standard deviation of response times as model parameters in the optimal transition models.}
\end{figure}

\newpage
\section{Supplementary F: Timing Uncertainty Task}

@@

    \caption*{\boldblue{Supplementary Figure 8: Timing Uncertainty Experiment}. One of two potential targets (dashed white rings; not visible to participants) would appear on the screen. Once a target appeared (solid white ring), participants were instructed to rapidly move their cursor to hit the displayed target. We used the mean and standard deviation of response times as model parameters in the optimal transition models.}
\end{figure}

\newpage
\section{Supplementary G: Guess Decisions}

@@

    \caption*{\boldblue{Supplementary Figure 9: Guess Decision Percentages for Experiment 1}. 
    Estimated percentage of guess decision percentages (y-axis) for each experimental condition (x-axis). During \boldblue{Experiment 1} participants were rewarded for successful trials and experienced a time constraint. Thus, we would expect participants to have faster response times in \boldblue{Experiment 1} than in the Response Time task \autocite*{milsteinInfluenceExpectedValue2007,shadmehrMovementVigorReflection2019a}. To account for this, we adjusted participant response times by subtracting 25ms from their median response time in the Response Time task. A trial was a guess decision if the difference between the stimulus movement onset and participant movement onset was less than that participant’s adjusted response time. We found that participants guessed significantly more often as the stimulus’s mean movement onset approached the time constraint.}
\end{figure}
